Vagueness, Noise, and Signaling

### 2015-08-13 paper/version_04/paper.tex
edits section 3

\newcommand{\myred}[1]{\textcolor{red}{#1}}

\doublespacing

@@

Signaling agents can adapt their dispositions to act, given some feedback about their past
success, in multiple ways. Usually, we would assume that changes in behavior should, at least
on average, lean towards increasing chances of communicative success. Such behavioral
adaptations can be described at different levels of abstraction. At the level of individual
agents, we can picture a more or less idealized process of how each agent adapts dispositions
for future actions based on various pieces of information available to the agent, e.g., about
the game, the other players' behavior and the like. More abstractly, at the level of a
population of agents, we can describe how average behavioral dispositions will evolve. The
population-level perspective abstracts over petty stochastic fluctuations and zooms in on the
general tendency or direction of evolution that ensues from some (often: simplified) behavior
at the agent-level.

To better understand the interaction of confusability of stimuli and selective pressure towards
successful communication, we look at a population-level dynamic that describes the most likely
evolutionary path of a population of signaling agents who imitate other agents' behavior but
may occasionally confuse states for one another. We call our dynamic \emph{imprecise
  conditional imitation dynamic} (\icid), because it derives from an update protocol based on
conditional imitation and allows for imprecise perception and realization of states. In the
special limiting case where imprecision vanishes, the process is just the well-known replicator
dynamic \citep{TaylorJonker1978:Evolutionary-St}. To appreciate the \icid and its motivation,
we first briefly review the replicator dynamic and its derivation from conditional imitation of
successes.

@@

The \emph{expected utility} of choices at each choice point is:
\begin{align*}
  \EU(\messg, \state, \Rstrat) & = \sum_{\act \in \Acts}
  \Rstrat(\act \probbar \messg) \cdot U_\sen(\state, \messg, \act) \\
  \EU(\act, \messg, \Sstrat) & = \sum_{\state \in
    \States} \frac{\Pr(\state) \cdot \Sstrat(\messg \probbar \state)}{\sum_{\state'} \Pr(\state') \cdot \Sstrat(\messg \probbar \state')} \cdot
  U_\rec(\state, \messg, \act) \,.
\end{align*}
Expected utilities are sums, over all possible concrete outcomes, of the utilities of these
outcomes, weighted by how likely they are to occur. For instance, the expected utility for the
sender of choosing message $\messg \in \Messgs$, given tate $\state \in \States$ and a receiver
strategy $\Rstrat$ is the sum, for each act $\act \in \Acts$, of the probability that the
receiver will choose act $\act$ given message $\messg$ times the actual utility of using act
$\act$, given $\state$ and $\messg$.

@@

For numerical simulations, it can be more convenient to work with a discrete-time formulation,
which maps current strategies $\Sstrat$ and $\Rstrat$ to strategies $\Sstrat'$ and $\Rstrat'$
that describe the population at the next discrete time step. Like the continuous-time version,
the discrete-time replicator dynamics tracks changes in frequency of choices as proportional to
their expected utilities \citep[e.g.][]{HofbauerSigmund1998:Evolutionary-Ga}:\footnote{The
  formulation given in Equation~(\ref{eq:RDDiscrete}) is adequate only for cases like the one
  we will be looking at in Section~\ref{sec:exploring-rdd}, where utilities are always
  non-negative and expected utilities are always positive.}\textsuperscript{,}\footnote{The
  notation $\propto$, for ``proportional to'', means that the right-hand side must possibly yet
  be normalized so as to have probabilities sum to one. In general, writing $P(x) \propto f(x)$
  for any function $f \mycolon X \rightarrow \mathds{R}$ is shorthand for
  $P(x) = \frac{f(x)}{\sum_{x' \in X} f(x')}$. The implicit normalizing denominators for the
  replicator equations are average expected utilities, e.g., for the sender:
  $\sum_{\messg'} \Sstrat(\messg' \probbar \state) \cdot \EU(\messg', \state,\Rstrat)$.}
\begin{align}
  \label{eq:RDDiscrete}
  \Sstrat'(\messg \probbar \state) & \propto \Sstrat(\messg \probbar \state) \cdot
    \EU(\messg, \state,\Rstrat)\,, & \Rstrat'(\act \probbar \messg) & \propto \Rstrat(\act \probbar \messg) \cdot
    \EU(\act, \messg,\Sstrat)\,.   
\end{align}
For a given choice point, say a state $\state$, the probability of seeing $\messg$ played by an
average agent in the population \emph{after} the update is proportional to the probability of
seeing it \emph{before} the update, which is $\Sstrat(\messg \probbar \state)$, times the
expected utility of $\messg$ at state $\state$. Intuitively put, frequencies of choices change
by a gradient of current frequency and a measure of how good they are.

@@

The replicator dynamic is an abstract population-level dynamic that describes the mean expected
change of behavioral dispositions in a population of signalers. There are several ways of
deriving the replicator dynamic from agent-level processes of behavioral adaptation. We focus
here on one of the simplest: \emph{conditional imitation of success} \citep[see][for this and
many other derivations]{Sandholm2010:Population-Game}. The main idea is that agents imitate the
behavior of another agent at some choice point with a probability that is proportional to the
expected utility of the latter agent's act for that choice point. Since the sim-max games that
we will be looking at in Section~\ref{sec:exploring-rdd} have positive utilities, we may
identify the switching probability with the expected utility.

Let's consider sender strategies, as the receiver case is parallel. Revision by imitation of
success proceeds as follows. Call (misleadingly!) the agent who gets a chance to change
behavior ``learner'' and the possibly to-be-imitated agent ``teacher.'' A random learner is
drawn from the population and given a chance to change behavior at choice point $\state$. The
probability that our learner plays $\messg$ is $\Sstrat(\messg \probbar \state)$. The learner
observes what a randomly sampled teacher does at $\state$. That would be $\messg'$ with
probability $\Sstrat(\messg' \probbar \state)$. The learner then starts using $\messg'$ instead
of $\messg$ with probability $\EU(\messg', \state, \Rstrat)$. (Of course, $\messg'$ and
$\messg$ could be the same.)

@@

\subsection{Imitation of success with noise}
\label{sec:imit-succ-with}

The above derivation of the replicator dynamic assumes that agents can discriminate choices and
choice points perfectly. Let's dispense with that assumption. With an eye towards sim-max
games, we will assume that receivers choose interpretation actions $\Acts = \States$ and that
any two states, but not messages, may be confused for one another.\footnote{It is relatively
  straightforward to also incorporate confusability of messages, but this is irrelevant to our
  present purposes.} Confusability of states will affect how agents behave (on average), how
they perceive others that they might want to imitate, and how good any behavioral disposition
actually is.

To keep matters simple, let us assume that agents carry pure dispositions to act, but noise can
affect their realization. As a sender, every agent maps states to messages: these are
subjectively perceived states, and no longer necessarily also the actually occurring state. As
a receiver, every agent maps messages to state interpretations: these are intended
interpretations that need not always be faithfully realized. This means that behavioral
strategies $\Sstrat \in (\Delta(\Messgs))^\States$ and $\Rstrat \in (\Delta(\States))^\Messgs$
represent the average proportions of \emph{actual} behavioral dispositions in the population,
the realization and observation of which can be distorted by confusion of similar states.

If $\mystate{a}$ is the actual state, let $P_o(\mystate{o} \probbar \mystate{a})$ be the
probability that a given agent observes state $\mystate{o}$. Similarly, if a given receiver
intends to select interpretation $\mystate{i}$, we will assume that the probability with which
state $\mystate{r}$ is realized is $P_r(\mystate{r} \probbar
\mystate{i})$.
Section~\ref{sec:exploring-rdd} gives a concrete specification of $P_o$ and $P_r$ in terms of
similarity of states. For the time being, it is best to treat $P_o$ and $P_r$ as arbitrary
conditional confusion probabilities.

@@

Expected utilities of choices at choice points should also take into account that actual states
need not be observed states, and intended interpretations need not be realized
interpretations. First, note that
\begin{align*}
  P_{\overline{o}}(\mystate{a} \probbar \mystate{o}) \propto P_a(\mystate{a}) P_o(\mystate{o}
    \probbar \mystate{a})
\end{align*}
is the probability that $\mystate{a}$ is actual if $\mystate{o}$ is observed. The probability
that a random sender produces $\messg$ when the actual state is $\mystate{a}$ is:
\begin{align*}
  P_{\Sstrat}(\messg \probbar \mystate{a}) = \sum_{\mystate{o}} P_o(\mystate{o} \probbar
  \mystate{a}) \Sstrat(\messg \probbar \mystate{o})\,. 
\end{align*}
The probability that the actual state is $\mystate{a}$ if a random sender produced $\messg$ is:
\begin{align*}
  P_{\overline{\Sstrat}}(\mystate{a} \probbar \messg) \propto P_a(\mystate{a})
  P_{\Sstrat}(\messg \probbar \mystate{a})\,. 
\end{align*}
The probability that $\mystate{r}$ is realized by a random receiver in response to
message $\messg$ is:
\begin{align*}
  P_{\Rstrat}(\mystate{r} \probbar \messg) = \sum_{\mystate{i}} P_r(\mystate{r} \ \probbar
  \mystate{i}) \Rstrat(\mystate{i} \probbar \messg)\,.
\end{align*}
This lets us capture the expected utilities for observed states (sender) and intended
interpretations (receiver) by taking into consideration what the likely actual states and
realized interpretations will be:
\begin{align*}
  \EU(\messg , \mystate{o}, \Rstrat) & = \sum_{\mystate{a}}  P_{\overline{o}}(\mystate{a}
  \probbar \mystate{o}) \sum_{\mystate{r}}  P_\Rstrat(\mystate{r} \probbar
  \messg) \utils(\mystate{a}, \messg, \mystate{r})\,, \\
  \EU(\mystate{i} , \messg, \Sstrat) & = \sum_{\mystate{a}}
  P_{\overline{\Sstrat}}(\mystate{a} \probbar \messg) \sum_{\mystate{r}}
  P_{r}(\mystate{r} \probbar \mystate{i})  \utils(\mystate{a}, \messg, \mystate{r})\,.
\end{align*}
If the conditional probabilities $P_o$ and $P_r$ are trivial, i.e., assign probability 0 to the
observability and realizability of non-identical states, above definitions reduce to the
previous definitions of expected utilities. This also legitimates the overload of notation.

Presence of potential imprecision in the form of non-trivial $P_o$ and $P_r$ will also affect
the dynamic that ensues from imitation of successes. Since imprecision works differently on
senders and receivers (the former confuse choice points, the latter confuse choices), we need
to look separately at each case.

@@

The continuous-time formulations in Equations~(\ref{eq:nRDSender}) and (\ref{eq:nRDReceiver})
have elegant and practical discrete-time solutions in:\footnote{To see how the discrete-time
  formulation gives rise to the continuous-time formulation, let's assume that update steps are
  infinitesimally small, so that
  \begin{align*}
    \dot{\sigma}(\messg \probbar \state) & = \sigma'(\messg \probbar \state) - \sigma(\messg
    \probbar \state)\\ 
    &= \frac{P_o(\messg \probbar \state) \myts \EU(\messg, \state, \Rstrat)
      - \Sstrat(\messg \probbar \state) \sum_{\messg'} P_o(\messg' \probbar \state) \myts
      \EU(\messg', \state, \Rstrat) }{ \sum_{\messg'} P_o(\messg' \probbar \state) \myts
      \EU(\messg', \state, \Rstrat)}\,.
  \end{align*}
By dropping the denominator, which is constant for all $\messg$ for fixed
$\state$, we obtain the above continuous-time formulation.}
\begin{align}
  \label{eq:nRD}
  \Sstrat'(\messg \probbar \state) & \propto P_o(\messg \probbar \state) \myts \EU(\messg,
  \state, \Rstrat)\,, & \Rstrat'(\state \probbar \messg) & \propto P_o(\state \probbar \messg)
  \myts \EU(\state, \messg, \Sstrat)\,.
\end{align}
This looks very much like the discrete-time formulation of the standard replicator dynamic in
(\ref{eq:RDDiscrete}). A minor difference is that expected utility here takes stochastic
confusability of states into account. The main difference is that here we are not looking at,
say, the probability $\Sstrat(\messg \probbar \state)$ with which $\messg$ is be played in
response to $\state$ on average in the population, but at the probability $P_o(\messg \probbar
\state)$ of seeing one agent pick $\messg$ given that another agent has observed
$\state$. Similarly for the receiver, we are not looking at occurrence probability
$\Sstrat(\state \probbar \messg)$ but at the probability $P_o(\state \probbar \messg)$ with
which an agent who observes $\messg$ observes another agent play $\state$. It is easy to check
formally and intuitive to appreciate that this major difference vanishes if confusion
probabilities $P_o$ and $P_r$ are trivial, i.e., assigning zero probability to all pairs of
non-identical states.

>>>>

\newcommand{\myred}[1]{\textcolor{red}{#1}} 
\renewcommand{\Pr}{\ensuremath{P}}

% \doublespacing

@@

Signaling agents can adapt their dispositions to act, given some feedback about their past
success, in multiple ways. Usually, we would assume that changes in behavior should, at least
on average, lean towards increasing chances of communicative success. Such behavioral
adaptations can be described at different levels of abstraction. At the level of individual
agents, we can picture a more or less idealized process of how each agent adapts dispositions
for future actions based on various pieces of information available to the agent. More
abstractly, at the level of a population of agents, we can describe how average behavioral
dispositions will evolve. The population-level perspective abstracts over petty stochastic
fluctuations and zooms in on the general tendency or direction of evolution that ensues from
behavior at the agent-level.

To better understand the interaction of confusability of stimuli and selective pressure towards
successful communication, we look at a population-level dynamic that describes the most likely
evolutionary path of a population of signaling agents who imitate other agents' behavior but
are liable to confuse states for one another. In the special limiting case where state
confusability vanishes, the process is just the well-known replicator dynamic
\citep{TaylorJonker1978:Evolutionary-St}, which should therefore be briefly reviewed first.

@@

The \emph{expected utility} of choices at each choice point is:\footnote{The notation
  $\propto$, for ``proportional to'', that is used here and hereafter means that the right-hand
  side must possibly yet be normalized so as to have probabilities sum to one. In general,
  writing $P(x) \propto f(x)$ for any function $f \mycolon X \rightarrow \mathds{R}$ is
  shorthand for $P(x) = \frac{f(x)}{\sum_{x' \in X} f(x')}$. }
\begin{align*}
  \EU(\messg, \state, \Rstrat) & = \sum_{\act \in \Acts}
  \Rstrat(\act \probbar \messg) \cdot U_\sen(\state, \messg, \act) \\
  \EU(\act, \messg, \Sstrat) & = \sum_{\state \in
    \States} P(\state \probbar \messg)  \cdot
  U_\rec(\state, \messg, \act)\,, \text{\ \  where } &
P(\state \probbar \messg) & \propto \Pr(\state) \cdot \Sstrat(\messg \probbar \state) \,.
\end{align*}
Expected utilities are sums, over all possible concrete outcomes, of the utilities of these
outcomes, weighted by how likely they are to occur. For instance, the expected utility for the
sender of choosing message $\messg$, given tate $\state$ and a receiver
strategy $\Rstrat$ is the sum, for each act $\act$, of the probability that the
receiver will choose $\act$ given $\messg$ times the actual utility of triple
$\state$, $\messg$ and $\act$.

@@

For numerical simulations, it can be more convenient to work with a discrete-time formulation,
which maps current strategies $\Sstrat$ and $\Rstrat$ to strategies $\Sstrat'$ and $\Rstrat'$
that describe the population at the next discrete time step. Like the continuous-time version,
the discrete-time replicator dynamics tracks changes in frequency of choices as proportional to
their expected utilities \citep[e.g.][]{HofbauerSigmund1998:Evolutionary-Ga}:\footnote{The
  formulation given in Equation~(\ref{eq:RDDiscrete}) is adequate only for cases like the one
  we will be looking at in Section~\ref{sec:exploring-rdd}, where utilities are always
  non-negative and expected utilities are always positive.}
\begin{align}
  \label{eq:RDDiscrete}
  \Sstrat'(\messg \probbar \state) & \propto \Sstrat(\messg \probbar \state) \cdot
    \EU(\messg, \state,\Rstrat)\,, & \Rstrat'(\act \probbar \messg) & \propto \Rstrat(\act \probbar \messg) \cdot
    \EU(\act, \messg,\Sstrat)\,.   
\end{align}
For a given choice point, say a state $\state$, the probability of seeing $\messg$ played by an
average agent in the population \emph{after} the update is proportional to the probability of
seeing it \emph{before} the update, which is $\Sstrat(\messg \probbar \state)$, times the
expected utility of $\messg$ at state $\state$. Intuitively put, frequencies of choices change
by a gradient of current frequency and a measure of how good they are.

@@

The replicator dynamic is an abstract population-level dynamic that describes the mean expected
change of behavioral dispositions in a population of signalers. There are several ways of
deriving the replicator dynamic from agent-level processes of behavioral adaptation. We focus
here on one of the simplest: \emph{conditional imitation of success}
\citep[see][]{Sandholm2010:Population-Game}. The main idea is that agents imitate the behavior
of other agents at some choice point with a probability that is proportional to the expected
utility of the latter agents' choice. Since the sim-max games that we will be looking at in
Section~\ref{sec:exploring-rdd} have positive utilities, we can identify the switching
probability with the expected utility.

Let's consider sender strategies, as the receiver case is parallel. Call (misleadingly!) the
agent who gets a chance to change behavior ``learner'' and the possibly to-be-imitated agent
``teacher.'' A random learner is drawn from the population and given a chance to change
behavior at choice point $\state$. The probability that our learner plays $\messg$ is
$\Sstrat(\messg \probbar \state)$. The learner observes what a randomly sampled teacher does at
$\state$. That would be $\messg'$ with probability $\Sstrat(\messg' \probbar \state)$. The
learner then starts using $\messg'$ instead of $\messg$ with probability
$\EU(\messg', \state, \Rstrat)$. (Of course, $\messg'$ and $\messg$ could be the same.)

@@

\subsection{Imitation of success with imprecision}
\label{sec:imit-succ-with}

The above derivation of the replicator dynamic assumes that agents can discriminate choices and
choice points perfectly. Let's dispense with that assumption. With an eye towards sim-max
games, we will assume that receivers choose interpretation actions $\Acts = \States$ and that
any two states, but not messages, may be confused for one another.\footnote{It is relatively
  straightforward to also incorporate confusability of messages, but this is irrelevant to our
  present purposes.} Confusability of states will affect how agents behave, how they perceive
the behavior of others, and the expected utilities of behavioral dispositions.

To keep matters simple, let us assume that agents carry pure dispositions to act. Noise can
affect the realization of these strategies. As a sender, every agent maps states to messages:
these are subjectively perceived states, and no longer necessarily also the actually occurring
states. As a receiver, every agent maps messages to state interpretations: these are intended
interpretations that need not always be faithfully realized. This means that behavioral
strategies $\Sstrat$ and $\Rstrat$ represent the average proportions of \emph{actual}
behavioral dispositions in the population, the realization and observation of which can be
distorted by agents' confusion of similar states.

If $\mystate{a}$ is the actual state, let $P_o(\mystate{o} \probbar \mystate{a})$ be the
probability that a given agent observes state $\mystate{o}$. Similarly, if a given receiver
intends to select interpretation $\mystate{i}$, let $P_r(\mystate{r} \probbar \mystate{i})$ be
the probability with which state $\mystate{r}$ is realized.  Section~\ref{sec:exploring-rdd}
gives a concrete (parameterized) specification of $P_o$ and $P_r$ in terms of similarity of
states. For the time being, it is best to treat $P_o$ and $P_r$ as arbitrary conditional
confusion probabilities.

@@

Expected utilities of choices at choice points should likewise take into account that actual states
need not be observed states, and intended interpretations need not be realized
interpretations. First, note that
\begin{align*}
  P_{\overline{o}}(\mystate{a} \probbar \mystate{o}) \propto P_a(\mystate{a}) P_o(\mystate{o}
    \probbar \mystate{a})
\end{align*}
is the probability that $\mystate{a}$ is actual if $\mystate{o}$ is observed by an agent. The
probability that a random sender produces $\messg$ when the actual state is $\mystate{a}$ is:
\begin{align*}
  P_{\Sstrat}(\messg \probbar \mystate{a}) = \sum_{\mystate{o}} P_o(\mystate{o} \probbar
  \mystate{a}) \Sstrat(\messg \probbar \mystate{o})\,. 
\end{align*}
The probability that the actual state is $\mystate{a}$ if a random sender produced $\messg$ is:
\begin{align*}
  P_{\overline{\Sstrat}}(\mystate{a} \probbar \messg) \propto P_a(\mystate{a})
  P_{\Sstrat}(\messg \probbar \mystate{a})\,. 
\end{align*}
The probability that $\mystate{r}$ is realized by a random receiver in response to
message $\messg$ is:
\begin{align*}
  P_{\Rstrat}(\mystate{r} \probbar \messg) = \sum_{\mystate{i}} P_r(\mystate{r} \ \probbar
  \mystate{i}) \Rstrat(\mystate{i} \probbar \messg)\,.
\end{align*}
This lets us capture the expected utilities for observed states (sender) and intended
interpretations (receiver) by taking into consideration what the likely actual states and
realized interpretations will be:
\begin{align*}
  \EU(\messg , \mystate{o}, \Rstrat) & = \sum_{\mystate{a}}  P_{\overline{o}}(\mystate{a}
  \probbar \mystate{o}) \sum_{\mystate{r}}  P_\Rstrat(\mystate{r} \probbar
  \messg) \utils(\mystate{a}, \messg, \mystate{r})\,, \\
  \EU(\mystate{i} , \messg, \Sstrat) & = \sum_{\mystate{a}}
  P_{\overline{\Sstrat}}(\mystate{a} \probbar \messg) \sum_{\mystate{r}}
  P_{r}(\mystate{r} \probbar \mystate{i})  \utils(\mystate{a}, \messg, \mystate{r})\,.
\end{align*}
If the conditional probabilities $P_o$ and $P_r$ are trivial, i.e., assign probability 0 to the
confusability of non-identical states, above definitions reduce to the
previous definitions of expected utilities. This also legitimates the overload of notation.

Presence of potential imprecision in the form of non-trivial $P_o$ and $P_r$ will also affect
the dynamic that ensues from imitation of successes. Since imprecision works slightly
differently on senders and receivers (the former confuse choice points, the latter confuse
choices), we need to look separately at each case.

@@

The continuous-time formulations in Equations~(\ref{eq:nRDSender}) and (\ref{eq:nRDReceiver})
have elegant and practical discrete-time solutions in:\footnote{To see how the discrete-time
  formulations gives rise to the continuous-time formulations, let's assume that update steps are
  infinitesimally small, so that, for the sender case:
  \begin{align*}
    \dot{\sigma}(\messg \probbar \state) & = \sigma'(\messg \probbar \state) - \sigma(\messg
    \probbar \state)\\ 
    &= \frac{P_o(\messg \probbar \state) \myts \EU(\messg, \state, \Rstrat)
      - \Sstrat(\messg \probbar \state) \sum_{\messg'} P_o(\messg' \probbar \state) \myts
      \EU(\messg', \state, \Rstrat) }{ \sum_{\messg'} P_o(\messg' \probbar \state) \myts
      \EU(\messg', \state, \Rstrat)}\,.
  \end{align*}
By dropping the denominator, which is constant for all $\messg$ for fixed
$\state$, we obtain the above continuous-time formulation.}
\begin{align}
  \label{eq:nRD}
  \Sstrat'(\messg \probbar \state) & \propto P_o(\messg \probbar \state) \myts \EU(\messg,
  \state, \Rstrat)\,, & \Rstrat'(\state \probbar \messg) & \propto P_o(\state \probbar \messg)
  \myts \EU(\state, \messg, \Sstrat)\,.
\end{align}
This looks very much like the discrete-time formulation of the standard replicator dynamic in
(\ref{eq:RDDiscrete}). A minor difference is that expected utility here takes stochastic
confusability of states into account. The main difference is that here we are not looking at,
say, the probability $\Sstrat(\messg \probbar \state)$ with which $\messg$ is played in
response to $\state$ on average in the population, but at the probability $P_o(\messg \probbar
\state)$ of seeing one agent pick $\messg$ given that another agent has observed
$\state$. Similarly for the receiver, we are not looking at occurrence probability
$\Rstrat(\state \probbar \messg)$ but at the probability $P_o(\state \probbar \messg)$ with
which an agent who observes $\messg$ observes another agent play $\state$. It is easy to check
formally and intuitive to appreciate that this major difference vanishes if confusion
probabilities $P_o$ and $P_r$ are trivial, i.e., assigning zero probability to all pairs of
non-identical states. The noisy replicator dynamic in (\ref{eq:nRD}) conservatively extends the
classic case in (\ref{eq:RDDiscrete}).
